Evaluation of Cache-Oblivious Data Structures

### 2008-03-24 proposal/proposal.tex
Some more changes to the proposal.

% TODO: abstract?
%\begin{abstract}
%\end{abstract}

@@

\section{Introduction}
%[A fundamental part of theoretical computer science is the study of algorithms: formal descriptions of how computations may be performed. In order for algorithms to be able to process information efficiently, it must be stored in a systematic way, called a data structure. Algorithms and data structures therefore complement each other: while algorithms describe how a computer may process information, data structures describe how information is stored in a (model of a) computer's memory.]

A fundamental part of theoretical computer science is the study of algorithms (formal descriptions of how computations may be performed) and data structures (descriptions of how information is organized and stored in computers). Traditionally, algorithms have been evaluated in a simplified model of computation. In this model it is assumed that a computer executes an algorithm in discrete steps. At each step it performs one elementary operation (e.g. comparing two numbers, adding one to another, storing a value in memory, et cetera). Each elementary operation is performed within a constant time. In this model, both storing and retrieving data values in memory is considered to be an elementary operation.

@@

It should be clear by now that a modern computer system lacks a single, central memory storage. Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage. At each of these levels, the next storage type is both significantly slower and significantly larger than the previous type. Typically, the processor can directly manipulate data contained in its registers only. To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy. Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).

@@

\subsection{Cache-Oblivious Memory Model}
The external memory model was generalized and simplified into the cache-oblivious model \cite{prokop1999coa} which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory. Data is transferred between the two in aligned data blocks of size B. In this model (and unlike the external memory model) the values of parameters like M and B are not known to algorithms and data structures, so they cannot be referenced explicitly in their definition. Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem). Like the external memory model, the cache-oblivious model can be used to describe every pair of adjacent levels in the memory hierarchy.

It should be noted that the cache-oblivious model can also be considered similar to the hierarchical model discussed above. The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
% This makes it easier to design algorithms in the oblivious model, because in the HMM the algorithm has the responsibility to use the cache correctly.

The key advantage of the cache-oblivious model is that algorithms that perform optimally when evaluated in this model are implicitly tuned to all levels in the real memory hierarchy at once. Therefore, these algorithms might perform better than algorithms that are tuned to a specific level in the hierarchy only.

@@

@@

% [TODO: add more recent related work]

% [TODO: find empirical results if they exists, or note lack of empirical data]

>>>>

\begin{abstract}
In modern computer hardware architecture, memory is organized in a hierarchy consisting of several types of memory with different sizes and access times. The cache-oblivious model has been proposed reflect this reality more accurately than traditional models. A number of algorithms and data structures have been proposed that perfom optimally in this model. The goal of the proposed research project is to implement a selection of cache-oblivious set-like data structures, evaluate their performance (in terms of execution time and memory use) in a realistic environment, and compare their performance to traditional data structures.
\end{abstract}

@@

\section{Introduction}
A fundamental part of theoretical computer science is the study of algorithms (formal descriptions of how computations may be performed) and data structures (descriptions of how information is organized and stored in computers). Traditionally, algorithms have been evaluated in a simplified model of computation. In this model it is assumed that a computer executes an algorithm in discrete steps. At each step it performs one elementary operation (e.g. comparing two numbers, adding one to another, storing a value in memory, et cetera). Each elementary operation is performed within a constant time. In this model, both storing and retrieving data values in memory is considered to be an elementary operation.

@@

It should be clear by now that a modern computer system lacks a central memory storage with uniform performance characteristics. Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage. At each of these levels, the next storage type is both significantly slower and significantly larger than the previous type. Typically, the processor can directly manipulate data contained in its registers only. To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy. Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).

@@

We will call algorithms that are designed to minimize the number of transfers in a two-level memory hierarchy ``cache-aware'' (as opposed to traditional ``cache-unaware'' algorithms) or ``cache-concious''. These algorithms typically rely on knowledge of the block size to achieve optimal performance.

\subsection{Cache-Oblivious Memory Model}
The external memory model was generalized and simplified into the cache-oblivious model \cite{prokop1999coa} which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory. Data is transferred between the two in aligned data blocks of size B. In this model (and unlike the external memory model) the values of parameters like M and B are not known to algorithms and data structures, so they cannot be referenced explicitly in their definition. Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).

It should be noted that the cache-oblivious model can also be considered similar to the hierarchical model discussed above. The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache. The cache-oblivious model is more realistic in this respect.

Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration. The key advantage of this class of algorithms is that they are implicitly tuned to all levels in the memory hierarchy at once. It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.

@@

Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list. This means buckets have to resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).

@@

Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-oblivious priority deque. Their main result is that although the cache-oblivious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
